On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-09-14 manuscript/pnas_revision.tex
first paragraph
more revisions

\dropcap{W}hen measuring the size of ``big data'', one should
consider the structures that implicitly give rise to one's data and
the computational complexities of modeling schemes that appropriately
take such structures into account. Phylogeographic methods
\cite{lemey2009bayesian, lemey2014unifying, holbrook2021massive,
  holbrook2022viral} model large-scale viral transmission between
human populations as a function of the shared evolutionary history of
the viral population of interest.  Data take the form of dates,
locations and genome sequences associated to individual viral
samples. Spatiotemporal structure interfaces with network structure
given by the phylogeny, or family tree, describing the viruses'
collective history beginning with the most recent common ancestor.
While one cannot directly observe this history, one may statistically
reconstruct the phylogenetic tree by positing that changes in the
viral genome happen randomly at regular intervals, thereby capturing
the intuition that viral samples with more differences between their
(aligned) sequences should find themselves further apart on the family
tree.

@@

We turn to our probabilistic bounds on the singular values of randomly
generated rate matrices, \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.
Although interesting in its own right, this result leads to
consequences for the bounds in \eqref{eq:qetq:exp:decay} and
\eqref{eq:ass:error} when applied in \cref{thm:bride:of:frankinstein}.
Before proceeding, we briefly introduce further mathematical
preliminaries associated with the so-called sub-exponential random
variables.
\begin{definition}
  \label{def:sub:exp:RV}
  A random variable $X$ is called \textit{sub-exponential} if there
  exists some constant $K > 0$ for which its tails satisfy
\begin{equation}\label{e:def_subexp_RV}
\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
\end{equation}
In this case, the \textit{sub-exponential norm} of $X$ is defined by
\begin{equation}\label{e:sub-exp_norm}
\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
\end{equation}
The \textit{class of sub-exponential distributions} is denoted by
\begin{align*}
L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
\end{align*}
\end{definition}

@@

\begin{remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}. In general, there appears to be no
  significant departure from the idea that the proxy form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{remark}

>>>>

\dropcap{P}hylogeographic methods
\cite{lemey2009bayesian, lemey2014unifying, holbrook2021massive,
  holbrook2022viral} model large-scale viral transmission between
human populations as a function of the shared evolutionary history of
the viral population of interest.  Data take the form of dates,
locations and genome sequences associated to individual viral
samples. Spatiotemporal structure interfaces with network structure
given by the phylogeny, or family tree, describing the viruses'
collective history beginning with the most recent common ancestor.
While one cannot directly observe this history, one may statistically
reconstruct the phylogenetic tree by positing that changes in the
viral genome happen randomly at regular intervals, thereby capturing
the intuition that viral samples with more differences between their
(aligned) sequences should find themselves further apart on the family
tree.

@@

We turn to our probabilistic bounds on the singular values of randomly
generated rate matrices, \cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.}.
Although interesting in its own right, this result leads to
consequences for the bounds in \eqref{eq:qetq:exp:decay} and
\eqref{eq:ass:error} when applied in \cref{thm:bride:of:frankinstein}.
Before proceeding, we briefly introduce further mathematical
preliminaries associated with the so-called sub-exponential random
variables.  To avoid confusion, note that the following definition uses the term in the same way as, e.g., \cite{vershynin:2018}, but that other definitions that mean quite the opposite (i.e., heavier than exponential tails) appear in the literature \cite{goldie1998subexponential}.
\begin{definition}
  \label{def:sub:exp:RV}
  A random variable $X$ is called \textit{sub-exponential} if there
  exists some constant $K > 0$ for which its tails satisfy
\begin{equation}\label{e:def_subexp_RV}
\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
\end{equation}
In this case, the \textit{sub-exponential norm} of $X$ is defined by
\begin{equation}\label{e:sub-exp_norm}
\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
\end{equation}
The \textit{class of sub-exponential distributions} is denoted by
\begin{align*}
L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
\end{align*}
\end{definition}

@@

\begin{remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}.  The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (\cref{def:sub:exp:RV}).  In general, there appears to be no
  significant departure from the idea that the proxy form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{remark}

### 2023-09-14 manuscript/revision_letter.tex
more revisions

@@

>>>>

 \newtheorem{@definition}{\bf Definition}
\newenvironment{definition}{\begin{@definition}\rm}{\end{@definition}}

@@

\newcommand{\bbP}{{\mathbb P}}
\newcommand{\bbE}{{\mathbb E}}
\begin{reply}
	You are correct that the truncated Cauchy (at 0) and the folded Cauchy are the same, but this distribution is not sub-exponential as we have defined the term in Definition \ref{def:sub:exp:RV}:
	\begin{quote}
\begin{definition}
	\label{def:sub:exp:RV}
	A random variable $X$ is called \textit{sub-exponential} if there
	exists some constant $K > 0$ for which its tails satisfy
	\begin{equation}\label{e:def_subexp_RV}
		\bbP(|X|\geq t) \leq 2 \exp\{-t/K\}, \quad \forall t \geq 0.
	\end{equation}
	In this case, the \textit{sub-exponential norm} of $X$ is defined by
	\begin{equation}\label{e:sub-exp_norm}
		\|X\|_{\psi_1} = \inf\big\{s > 0: \bbE \exp(|X|/s) \leq 2\big\}.
	\end{equation}
	The \textit{class of sub-exponential distributions} is denoted by
	\begin{align*}
		L_{\psi_1} = \big\{F_X(dx): \|X\|_{\psi_1} < \infty \big\}.
	\end{align*}
\end{definition}
	\end{quote}
Here, sub-exponential means the opposite of heavy-tailed.  The folded Cauchy distribution represents a distinct failure of the sub-exponential assumption.  In order to avoid further confusion, we have added the following sentence prior to Definition 1:
\begin{quote}
	To avoid confusion, note that the following definition uses the term in the same way as, e.g., \cite{vershynin:2018}, but that other definitions that mean quite the opposite (i.e., heavier than exponential tails) appear in the literature \citep{goldie1998subexponential}.
\end{quote}
Additionally, we have added the following sentence to Remark 4:
\begin{quote}
	The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (Definition \ref{def:sub:exp:RV}).
\end{quote}
\end{reply}
